High-Dimensional Regularized Discriminant Analysis

### 2013-12-29 reports/paper-hdrda.Rnw
The model-selection section is described in more detail. Needs a little more polish before closing out #11

@@

A primary contributing factor to the efficiency of Algorithm
\ref{alg:hdrda_model_selection} is our usage of the compact singular value
decomposition (SVD) (sometimes called the reduced or fast SVD), which we
describe here. Rather than computing the eigenvalue decomposition of
$\widehat{\bm \Sigma}$ to obtain $\bm U_1$, we instead obtain $\bm U_1$ by
computing the eigendecomposition of a much smaller $N \times N$ matrix when $p
\gg N$ \citep{Hastie:2008dt}. Recall that the pooled sample covariance matrix is
$\widehat{\bm \Sigma} = \bm X^{T} \bm X$\footnote{TODO: $1/N$ is missing. Add
  this. It will be absorbed into the eigenvalues.}  Then, because $\bm X = \bm M
\bm D^{1/2} \bm U^{T}$ by the SVD, we have the eigendecomposition $\widehat{\bm
  \Sigma} = \bm U \bm D \bm U^{T}$, where $\bm M \in \mathbb{R}_{N \times N}$ is
orthogonal, $\bm D^{1/2} \in \mathbb{R}_{N \times N}^{\ge}$ is a diagonal matrix
consisting of the singular values of $\bm X$, and $\bm U \in \mathbb{R}_{p
  \times N}$ is orthogonal. Now, $\bm X \bm X^{T} = \bm M \bm D \bm M^{T}$ is
the eigenvalue decomposition of $\bm X \bm X^{T}$, where $\bm D$ is the matrix
of eigenvalues of $\bm X \bm X^{T}$ and the columns of $\bm M$ are the
corresponding eigenvectors. Hence, we obtain $\bm M$ and $\bm D$ from an
eigendecomposition of $\bm X \bm X^{T} \in \mathbb{R}_{N \times N}$. Next, we
compute $\bm U = \bm X^{T} \bm M \bm D^{+/2}$, where $\bm D^{+/2} = \bm
D_q^{-1/2} \oplus \bm 0_{N-q}$. We then determine $q$, the number of numerically
nonzero eigenvalues present in $\bm D$, by calculating the number of eigenvalues
that exceeds some tolerance value, say, $\epsilon = 1 \times 10^{-6}$. After
calculating $q$, we extract $\bm U_1$ by retaining only the first $q$ columns of
$\bm U$.

% TODO: Discuss things that need only be computed once
% TODO: Discuss operations independent of tuning parameters

\subsection{Computational Complexity}

Conveniently, we see that the inverse and determinant of $\bm W_k$ are each in
terms of the inverse and determinant of $\bm Q_k \in \mathbb{R}_{n_k \times
  n_k}$, respectively, as well as the diagonal matrix $\bm \Gamma_k \in
\mathbb{R}_{q \times q}$. Hence, the resulting formulation requires $O(n_k p q)$
calculations to compute both $\bm W_k^{-1}$ and $| \bm W_k |$. The expressions
given in Proposition \ref{proposition:hdrda-W_k} also expedite the selection of
$\lambda$ and $\gamma$ via cross-validation because the most time-consuming
matrix operation involved in computing $\bm W_k^{-1}$ and $|\bm W_k|$ is $\bm
X_k \bm U_1 \in \mathbb{R}_{n_k \times q}$, which is independent of $\lambda$
and $\gamma$. Therefore, we need calculate $\bm X_k \bm U_1$ only once for each
cross-validation fold. The subsequent operations in calculating $\bm W_k^{-1}$
and $|\bm W_K|$ can be simply updated for different pairs of $\lambda$ and
$\gamma$.

\begin{enumerate}
  \item $\bm W_k^{-1}$\footnote{TODO: Need complexity for the following}
  \item $|\bm W_k|$
  \item (12)
\end{enumerate}

@@

We used version 3.0.2 of the open-source statistical software {\tt R} for the
classification study. We used the implementations of the Witten and Guo
classifiers from the {\tt penalizedLDA} and {\tt rda} packages, respectively, on
CRAN. Because the {\tt rda} package does not perform the authors' ``Min-Min''
rule automatically, we applied this rule in our {\tt R} code. We also utilized
our implementations of the Dudoit, Pang, and Tong classifiers available in the
{\tt sparsediscrim} package. We used the default settings for each classifier
with the exception that we explicitly set prior probabilities as equal, if the
option was available. For the \emph{HDRDA} classifier in
\eqref{eq:hdrda-decomposed}, we examine the classification performance of two
models. For the first \emph{HDRDA} model, we set $\alpha_k = 1$, $k = 1, \ldots,
K$, so that the covariance-matrix estimator \eqref{eq:hdrda-cov} employed
resembled \eqref{eq:ridge-estimator}. We estimated $\lambda$ from a grid of 21
equidistant candidate values between 0 and 1, inclusively. Similarly, we
estimated $\gamma$ from a grid consisting of the values $10^{-1}, \ldots, 10^4$,
and $10^5$. We selected optimal estimates of $\lambda$ and $\gamma$ using
10-fold cross-validation \citep{Hastie:2008dt}. For the second model, we set
$\alpha_k = 1 - \gamma$, $k = 1, \ldots, K$, to resemble
\citeauthor{Friedman:1989tm}'s parameterization. We estimated both $\lambda$ and
$\gamma$ from a grid of 21 equidistant candidate values between 0 and 1,
inclusively. Each of the microarray data sets listed below can be found in
\ref{suppA}.

>>>>

\bibpunct{(}{)}{;}{a}{,}{,}

@@

A primary contributing factor to the efficiency of Algorithm
\ref{alg:hdrda_model_selection} is our usage of the compact singular value
decomposition (SVD) (sometimes called the reduced or fast SVD), which we
describe here. Rather than computing the eigenvalue decomposition of
$\widehat{\bm \Sigma}$ to obtain $\bm U_1$, we instead obtain $\bm U_1$ by
computing the eigendecomposition of a much smaller $N \times N$ matrix when $p
\gg N$ \cite[Chapter~18.3.5]{Hastie:2008dt}. Recall that the pooled sample
covariance matrix is $\widehat{\bm \Sigma} = \bm X^{T} \bm X$\footnote{TODO:
  $1/N$ is missing. Add this. It will be absorbed into the eigenvalues.}  Then,
because $\bm X = \bm M \bm D^{1/2} \bm U^{T}$ by the SVD, we have the
eigendecomposition $\widehat{\bm \Sigma} = \bm U \bm D \bm U^{T}$, where $\bm M
\in \mathbb{R}_{N \times N}$ is orthogonal, $\bm D^{1/2} \in \mathbb{R}_{N
  \times N}^{\ge}$ is a diagonal matrix consisting of the singular values of
$\bm X$, and $\bm U \in \mathbb{R}_{p \times N}$ is orthogonal. Now, $\bm X \bm
X^{T} = \bm M \bm D \bm M^{T}$ is the eigenvalue decomposition of $\bm X \bm
X^{T}$, where $\bm D$ is the matrix of eigenvalues of $\bm X \bm X^{T}$ and the
columns of $\bm M$ are the corresponding eigenvectors. Hence, we obtain $\bm M$
and $\bm D$ from an eigendecomposition of $\bm X \bm X^{T} \in \mathbb{R}_{N
  \times N}$. Next, we compute $\bm U = \bm X^{T} \bm M \bm D^{+/2}$, where $\bm
D^{+/2} = \bm D_q^{-1/2} \oplus \bm 0_{N-q}$. We then determine $q$, the number
of numerically nonzero eigenvalues present in $\bm D$, by calculating the number
of eigenvalues that exceeds some tolerance value, say, $\epsilon = 1 \times
10^{-6}$. After calculating $q$, we extract $\bm U_1$ by retaining only the
first $q$ columns of $\bm U$.

As a result of the compact SVD, we need calculate $\bm X_c \bm U_1$ only once
per cross-validation fold, requiring $O(p q N) \approx (p N^2)$ calculations.
Hence, the computational costs of expensive calculations, such as matrix
inverses and determinants, are greatly reduced because they are performed in the
$q$-dimensional subspace. Similarly, we reduce the dimension of the test data
set by calculating $\bm X_c \bm U_1$ once per fold. Conveniently, we see that
the most costly computation involved in $\bm Q_k$ and $\bm W_k^{-1}$ is $\bm X_k
\bm U_1$, which can be extracted from $\bm X_c \bm U_1$. Thus, after the initial
calculation of $\bm X_c \bm U_1$ per cross-validation fold, $\bm Q_k$ requires
$O(TODO)$ operations. Because $\bm Q_k \in \mathbb{R}_{n_k \times n_k}$, both
its determinant and inverse requires $O(n_k^3)$ operations. This is especially
advantageous when $K$ is large, such that $n_k$ is much smaller than $N$. Thus,
$\bm W_k^{-1}$ requires $O(TODO)$ operations Also, the inverse and determinant
of the diagonal matrix $\bm \Gamma_k \in \mathbb{R}_{q \times q}$ requires
$O(q)$ operations. Hence, the $| \bm W_k |$ requires $O(q n_k^3)$ operations,
which is efficient when $p >> N$.

The expressions given in Proposition \ref{proposition:hdrda-W_k} also expedite
the selection of $\lambda$ and $\gamma$ via cross-validation because the most
time-consuming matrix operation involved in computing $\bm W_k^{-1}$ and $|\bm
W_k|$ is $\bm X_k \bm U_1 \in \mathbb{R}_{n_k \times q}$, which is independent
of $\lambda$ and $\gamma$. The subsequent operations in calculating $\bm
W_k^{-1}$ and $|\bm W_K|$ can be simply updated for different pairs of $\lambda$
and $\gamma$.\footnote{TODO: Need to clean up this paragraph so that it flows
  from above. Also, elaborate.}

@@

We used version 3.0.2 of the open-source statistical software {\tt R} for the
classification study. We used the implementations of the Witten and Guo
classifiers from the {\tt penalizedLDA} and {\tt rda} packages, respectively, on
CRAN. Because the {\tt rda} package does not perform the authors' ``Min-Min''
rule automatically, we applied this rule in our {\tt R} code. We also utilized
our implementations of the Dudoit, Pang, and Tong classifiers available in the
{\tt sparsediscrim} package. We used the default settings for each classifier
with the exception that we explicitly set prior probabilities as equal, if the
option was available. For the \emph{HDRDA} classifier in
\eqref{eq:hdrda-decomposed}, we examine the classification performance of two
models. For the first \emph{HDRDA} model, we set $\alpha_k = 1$, $k = 1, \ldots,
K$, so that the covariance-matrix estimator \eqref{eq:hdrda-cov} employed
resembled \eqref{eq:ridge-estimator}. We estimated $\lambda$ from a grid of 21
equidistant candidate values between 0 and 1, inclusively. Similarly, we
estimated $\gamma$ from a grid consisting of the values $10^{-1}, \ldots, 10^4$,
and $10^5$. We selected optimal estimates of $\lambda$ and $\gamma$ using
$V$-fold cross-validation with $V = 10$ \citep{Hastie:2008dt}. For the second
model, we set $\alpha_k = 1 - \gamma$, $k = 1, \ldots, K$, to resemble
\citeauthor{Friedman:1989tm}'s parameterization. We estimated both $\lambda$ and
$\gamma$ from a grid of 21 equidistant candidate values between 0 and 1,
inclusively. Each of the microarray data sets listed below can be found in
\ref{suppA}.
